Aggregated Product System Models in Life Cycle Assessment: Achieving Transparency and Privacy

### 2017-10-20 aggregation/agg-disclosure.tex
move basic agg back into the study definition

The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.  Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.  These three submatrices together are referred to as the product system model, or PSM.

This set of three submatrices can be used to represent the PSM; thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.  Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.  A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.  Thus, a formal disclosure of the PSM can be stated in six parts:
\begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
\item\label{itm:fg} An ordered list of foreground nodes (rows/columns of $A_f$);
\item\label{itm:bg} An ordered list of background flows, making reference to a particular background database (rows of $A_d$, mapping to columns of $B_x$);
\item\label{itm:em} An ordered list of exchanges with environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
\item\label{itm:af} A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
\item\label{itm:ad} a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
\item\label{itm:bf} a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
\end{enumerate}

A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
\begin{itemize}
  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references to $B_x$ and $\mathbf{e}$;
  \item The disclosure includes only information for which the author can claim \textbf{authority};
  \item The disclosure includes only information for which the study is the \textbf{primary} source.
\end{itemize}
As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{e}$.  These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.  However, the the challenges of operationalizing this requirement are tremendously complex.  Some implications of the requirement for shared agreement are considered in the Discussion.

>>>>

The foreground matrix $A_f$, dependency matrix $A_d$, and foreground emission matrix $B_f$ now contain all study-specific information.  Assuming that the appropriate background LCIDB $B_x$, and the vector of characterization factors $\mathbf{e}$, are available and held in common between the author and the reader, then a disclosure that precisely describes $A_f$, $A_d$ and $B_f$ is sufficient to reproduce the study result, as formulated in Eq.~\ref{eqn:study}.  These three submatrices together can be considered a complete representation of the PSM.

Thus, the problem of how to disclose an LCA study design is reduced to the problem of how to communicate the contents of these submatrices accurately.  Recognizing that all three matrices are likely to be sparse (that is, that most of the entries are zero), the most efficient disclosure would be to only report non-zero entries.  A sparse matrix can be represented by a list of 3-tuples indicating (row, column, value) for each non-zero entry.  Thus, a formal disclosure of the PSM can be stated in six parts:
\begin{enumerate}[label={\em d-\roman*}., ref={\em d-\roman*}]
\item\label{itm:fg} An ordered list of foreground nodes (rows/columns of $A_f$);
\item\label{itm:bg} An ordered list of background flows, making reference to a particular background database (rows of $A_d$, mapping to columns of $B_x$);
\item\label{itm:em} An ordered list of exchanges with environment (rows of $B_f$, mapping to entries in $\mathbf{e}$);
\item\label{itm:af} A list of 3-tuples for nonzero elements of $A_f$, in which the row and column are indices into Item~\ref{itm:fg};
\item\label{itm:ad} a list of 3-tuples for nonzero elements of $A_d$, in which the rows are indices into Item~\ref{itm:bg} and the columns are indices into Item~\ref{itm:fg};
\item\label{itm:bf} a list of 3-tuples for nonzero elements of $B_f$, in which the rows are indices into Item~\ref{itm:em} and the columns are indices into Item~\ref{itm:fg}.
\end{enumerate}

A disclosure containing these six elements satisfies most of the requirements established at the top of this section (the exception is ensuring the privacy of confidential data, which is treated later in the article):
\begin{itemize}
  \item The result is \textbf{computable} using Eq.~\ref{eqn:study};
  \item the computation is \textbf{complete} and \textbf{minimal} because all components are required and no superfluous components are included;
  \item The results are \textbf{reproducible} if the reader is able to accurately interpret the author's references in \ref{itm:bg} and \ref{itm:em};
  \item The disclosure includes only information for which the author can claim \textbf{authority};
  \item The disclosure includes only information for which the study is the \textbf{primary} source.
\end{itemize}
As mentioned, the proper interpretation of the disclosure requires that the reader accurately interprets the author's references to background databases and elementary flows, including mutually consistent implementation of the LCIA vector $\mathbf{e}$.  These aspects are out of scope for the purposes of this paper, and we will assume that a shared understanding between author and reader can be reached.  However, the the challenges of operationalizing this requirement are tremendously complex.  Some implications of the requirement for shared agreement are considered in the Discussion.

### 2017-10-20 aggregation/agg-intro.tex
move basic agg back into the study definition

Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver 
them to a consumer.  The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
Increasingly, the results of LCA studies are appearing in environmental declarations, corporate sustainability reports, and product marketing information.  LCA is also gaining prominence as a tool for developing and evaluating environmental policy.  
%
At the core of an LCA study is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.  Product system models encode information about how products are manufactured, distributed, used by the consumer, and what happens to them after they are disposed.  A product system can be divided into a foreground, which denotes the portions of the life cycle whose operations are directly modeled by the study, and a background, which represents the global industrial system \citep{SETAC_inventory_1998}.  While the foreground is often modeled from direct observation or other primary data, background processes are typically drawn from a life cycle inventory (LCI) database prepared by a third party.

Although there has been considerable effort to normalize and harmonize 
LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing product system models.  These models are often highly complex and include countless modeling decisions, approximations, and assumptions made by a study author \citep{Lloyd2007, reap2008_I}.  Consequently, while comparative evaluations can be made regarding alternative cases or scenarios within a single study, comparisons across studies are much more challenging \citep{Heath2012, Henriksson2014}.  Even when relatively simple and homogeneous systems are considered, wide variations can be found in the results of studies from different authors \citep{van_der_Harst_2013, Turconi_2013}.
A lack of transparency in reporting, particularly regarding definitions of study scope and system boundary,
is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.  A further challenge to transparency can be found in the use of input data that are confidential or proprietary \citep{Kuczenski_2017}.  One consequence of all these challenges is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.

@@

An LCA study result is ultimately an assertion that for some constructed product model, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.  But if the product model itself is ambiguously stated, then the significance of the result is ambiguous as well.  For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.  But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.  When a study is reviewed, it is often the written report and \emph{not} the product system model itself that is the object of review.  In the end, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the product model itself is not available.

A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the product system model.  Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the product system model and thus, no automatic mechanisms for interpretation or review of published LCA studies.

>>>>

Life cycle assessment (LCA) is a standard methodology to estimate the potential environmental impacts of products or services by modeling the network of industrial processes that must occur to deliver 
them to a consumer.  The technique is well-established, widely practiced, and draws on an extensive body of standardization and scholarship.
Increasingly, the results of LCA studies are appearing in environmental declarations, corporate sustainability reports, and product marketing information.  LCA is also gaining prominence as a tool for developing and evaluating environmental policy.  
%
At the core of an LCA study is a model of a ``product system,'' which represents a collection of ``processes $\ldots$ performing one or more defined functions'', that describes the life cycle of a product \citep{iso14044}.  The product system model, or PSM, encodes information about how a product is manufactured, distributed, used by the consumer, and what happens to it after it is disposed.  A product system can be divided into a foreground, which denotes the portions of the life cycle whose operations are directly modeled by the study, and a background, which represents the global industrial system \citep{SETAC_inventory_1998}.  While the foreground is often modeled from direct observation or other primary data, background processes are typically drawn from a life cycle inventory (LCI) database prepared by a third party.

Although there has been considerable effort to normalize and harmonize 
LCI database design methodology and data interchange \citep{UNEP_2011, JRC_ILCD_ELCD_2013, Mila_e_Canals_2015, Ingwersen_JLCA_2015}, less attention has been paid to harmonizing PSMs.  These models are often highly complex and include countless modeling decisions, approximations, and assumptions made by a study author \citep{Lloyd2007, reap2008_I}.  Consequently, while comparative evaluations can be made regarding alternative cases or scenarios within a single study, comparisons across studies are much more challenging \citep{Heath2012, Henriksson2014}.  Even when relatively simple and homogeneous systems are considered, wide variations can be found in the results of studies from different authors \citep{van_der_Harst_2013, Turconi_2013}.
A lack of transparency in reporting, particularly regarding definitions of study scope and system boundary,
is a crucial challenge to the interpretation of results in comparative analysis \citep{Cleary2009, Laurent_2014}.  A further challenge to transparency can be found in the use of input data that are confidential or proprietary \citep{Kuczenski_2017}.  One consequence of all these challenges is the high cost and uneven rigor of critical review, in which the complexities of LCA come face to face with the limitations of current practice \citep{Curran2014}.

@@

An LCA study result is ultimately an assertion that for some constructed PSM, the delivery of a particular reference flow is associated with a certain amount of environmental impact or potential impact.  But if the PSM itself is ambiguously stated, then the significance of the result is ambiguous as well.  For a computational model to have scientific significance, its results need to be both verifiable and replicable by other scientists \citep{Fomel_2009, Mesirov_2010}.  But in the case of LCA, the structure and contents of the model are never stated precisely; instead they are \emph{described} in a written report, and the description must be interpreted by a reader.  When a study is reviewed, it is often the written report and \emph{not} the PSM itself that is the object of review.  In the end, there is no current practice that permits a reviewer to verify the correctness of the computations inherent in an LCA study if access to the PSM itself is not available.

A common theme in all these controversies is the inability for authors and their readers to formally agree on the structure and content of the PSM.  Although the general structure for life cycle assessment can be stated formally, there is no current practice for formally describing the PSM and thus, no automatic mechanisms for interpretation or review of published LCA studies.

### 2017-10-20 aggregation/agg-lca-study.tex
Work on study review
move basic agg back into the study definition

The simplest conceivable foreground is simply what is given as the final demand $\mathbf{y}$, in other words the list of background processes invoked by the study and their activity levels.  For an LCA computation as expressed in Eq.~\ref{eqn:leontief}, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
\begin{equation}
  \begin{array}{cc}
s =& \mathbf{e}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
 =& \mathbf{e}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
  \end{array}
  \label{eqn:canonical}
\end{equation}
where $\tilde{A} = \left[\begin{smallmatrix}
 0 & 0 \\
 \mathbf{y} & A
  \end{smallmatrix}\right]$, $\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.  \textit{author's note: a proof for this equality is provided in the electronic supplementary information.}  This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.

Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.  For clarity in this paper, I will call the augmented $\tilde{A}$ and $\tilde{B}$ matrices an \emph{LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the \emph{canonical} functional unit.

In a canonical LCA foreground study, the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground.
An LCA foreground study can be written in block-triangular form:
\begin{eqnarray}
\tilde{A} &= \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right]\\
  \tilde{B} &= \left[\begin{array}{cc} B_f & B   \end{array}\right]
\label{eqn:foreground}
\end{eqnarray}

\input{fig_foreground}
In this case, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.  The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions. The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.  As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.\footnote{It should be noted that this condition does not hold for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells.  The approach presented here applies only to LCA studies in which the background system is invariant with respect to the foreground. This class includes the vast majority of LCA case studes, and very likely includes all case studies prepared using commercial LCA software.}

It is currently common practice for background database maintainers to pre-compute the life cycle inventory results for their databases.  Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, Eq.s~\ref{eqn:foreground} can be rewritten as:
\begin{equation}
 \tilde{A} = \left[\begin{array}{cc} 
A_f &  0 \\
A_d & 0 
   \end{array}\right];\;\;\;  \tilde{B} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
\label{eqn:fg_bg}
\end{equation}
The proof of the equivalency of Eq.s\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.  In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining there pertains to the study foreground.  Furthermore, the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported.

@@

>>>>

The simplest conceivable foreground is simply what is given as the final demand $\mathbf{y}$, in other words the list of background processes invoked by the study and their activity levels.  For an LCA computation as expressed in Eq.~\ref{eqn:leontief}, the same result can be obtained by relocating the $\mathbf{y}$ vector into an augmented $A$ matrix, augmenting the $B$ matrix with a vector of zeros, and replacing the final demand with a vector $\tilde{\mathbf{y}}$ in which the first entry is 1 and all successive entries are 0:
\begin{equation}
  \begin{array}{cc}
s =& \mathbf{e}^T\times B\times\left(I-A\right)^{-1}\times\mathbf{y} \\
 =& \mathbf{e}^{T}\times\tilde{B}\times\left(I-\tilde{A}\right)^{-1}\times\tilde{\mathbf{y}}
  \end{array}
  \label{eqn:canonical}
\end{equation}
where $\tilde{A} = \left[\begin{smallmatrix}
 0 & \mathbf{0} \\
 \mathbf{y} & A
  \end{smallmatrix}\right]$, $\tilde{B} = [ \mathbf{0},\, B ]$, and $\tilde{\mathbf{y}} =  [ 1,\, 0,\, 0 ,\,\ldots,\, 0]^{T}$.
It can easily be shown that Eq.~\ref{eqn:canonical} holds for any $\mathbf{y}$.  \textit{author's note: a proof for this equality is provided in the electronic supplementary information.}  This is intuitively true because the $\tilde{\mathbf{y}}$ selects the first column of the augmented $\tilde{A}$ matrix as the ``final demand,'' but the ``direct requirements'' of that column are precisely the final demand from the prior, un-augmented $A$ matrix.

Of course, most studies have a more elaborate foreground than simply the demand for processes in the background database, and in these cases the $A$ and $B$ matrices can be further augmented with additional foreground content.  An LCA study augmented with foreground information can be written in block-triangular form:
\begin{equation}
\tilde{A} = \left[\begin{array}{cc}
A_f & 0 \\
A_d &  A
  \end{array}
\right];\;\;\;
  \tilde{B} = \left[\begin{array}{cc} B_f & B   \end{array}\right]
\label{eqn:foreground}
\end{equation}
Here, the submatrix $A_f$ represents the foreground (the first column of which delivers the functional unit of the study); $A$ represents the background, and the rectangular matrix $A_d$ represents the dependency of the foreground on the background.  The ordered $\tilde{B}$ matrix is similarly partitioned into $B_f$, which includes foreground emissions, and $B$ which includes background emissions. The constitutive characteristic of this formulation is that the submatrix in the top right corner of $\tilde{A}$ is zero.  As long as this is the case, all computations regarding the background are invariant with respect to any foreground, and they can be computed in advance.
\input{fig_foreground}

For clarity, I will call the augmented $\tilde{A}$ and $\tilde{B}$, in which the foreground content always appears to the left of the background content; the first column delivers the study's functional unit; and the background does not depend on the foreground; to be a \emph{canonical LCA foreground study}, for which $\tilde{\mathbf{y}}$ is the canonical functional unit.  This formulation can be applied to the vast majority of LCA case studes, and very likely to all case studies prepared using commercial LCA software.  However, it does not apply for so-called ``integrated hybrid'' models, in which all quadrants contain non-zero cells (for instance, \cite{Gibon_2016}).

It is currently common practice for background database maintainers to pre-compute the life cycle inventory results for their databases.  Using the notation $B_x = B \times (I - A)^{-1}$ to represent the pre-computed background LCI database, Eq.s~\ref{eqn:foreground} can be rewritten as:
\begin{equation}
 \tilde{A} = \left[\begin{array}{cc} 
A_f &  0 \\
A_d & 0 
   \end{array}\right];\;\;\;  \tilde{B} = \left[\begin{array}{cc} B_f & B_x \end{array}\right ]
\label{eqn:fg_bg}
\end{equation}
The proof of the equivalency of Eq.s~\ref{eqn:foreground} and \ref{eqn:fg_bg} is provided in the supplementary information to \cite{Kuczenski_JLCA_2015}.  In this form, the main content of the LCIDB has been removed from the $\tilde{A}$ matrix, and the only information remaining there pertains to the study foreground.  Furthermore, the final demand vector has been replaced with a canonical functional unit and no longer needs to be explicitly reported.

@@

The activity levels of the foreground nodes are represented by $\tilde{\mathbf{x}}$, which is the result of $\tilde{\mathbf{y}}$ selecting the first column of $(I-A_f)^{-1}$.
%% An important intermediate result is the vector of activity levels for the foreground nodes $\tilde{\mathbf{x}}$, also visualized in Figure~\ref{fig:foreground}:
%% \begin{equation}
%% \tilde{\mathbf{x}} = (I - A_f)^{-1} \times \tilde{\mathbf{y}}
%% \label{eqn:inv}
%% \end{equation}
This vector can be used to describe an \emph{aggregated foreground}:
\begin{equation}
\begin{array}{rl}
    \tilde{\mathbf{b}}_f & = B_f \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{a}}_d & = A_d \times \tilde{\mathbf{x}} \\
    \tilde{\mathbf{b}}_x & = B_x \times \tilde{\mathbf{a}}_d
\end{array}
\label{eqn:agg}
\end{equation}
The study result can also be computed from the aggregated foreground:
\begin{equation}
\begin{array}{rl}
s & = \mathbf{e}^T \times( B_f + B_x \times A_d) \times \tilde{\mathbf{x}} \\
  & = \mathbf{e}^T \times( \tilde{\mathbf{b}}_f + B_x \times\tilde{\mathbf{a}}_d) \\
  & = \mathbf{e}^T \times( \tilde{\mathbf{b}}_f + \tilde{\mathbf{b}}_x) \\
\end{array}
\end{equation}
